TIMES Ireland Model: Residential Sector

### 2021-05-21 manuscript.tex
Send 1st draft to Fionn, Hannah and Olex

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}
\label{Intro:Climate Targerts}
\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
The EU's updated Nationally Determined Contribution (NDC), increases EU's GHG reduction target from 40\% (first NDC) to at least 55\% (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emissions target would change from 30\% to 45\% reduction in 2030 compared to 2005. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

@@

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports. Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for some of these barriers. To put Irelands targets in perspective the ``Leaders' Summit on Climate" in April 2021, brought together the largest GHG emitters to update and raise ambition of their 2030 climate targets. Comparing these new targets with the same base year of 2018, Fig \ref{fig:GHG2030AGR} shows Ireland's national climate target is ambitious, especially when the share of agriculture GHG emissions is considered. Ireland has some distinct barriers and ambitious targets, the Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets, one of those modelling tools is discussed in the next section.

@@

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
A Energy System Optimisation Model (ESOM) can take account of a large number of dynamic variables, and complex interconnections in each time period and each economic sector and determine a least cost solution. The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA).  TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
\par While other ESOM frameworks are available to explore Ireland's decarbonisation pathway such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. The expertise among energy policy researchers in Ireland and the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. \par Previous versions of the Irish TIMES model were extracted from the PET36 model (Pan European TIMES model), and then updated with specific national data and assumptions, these models were used to provide informative reports to Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}. A new TIMES model called ``TIMES Ireland Model" (TIM) was built in 2021, TIM is structured to explore Ireland's new climate targets. 
TIMES has also been widely used to internationally, with some studies focusing on the residential sector. For example, Li studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed how higher thermal efficiency in buildings might hamper co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}.
TIMES has been used in Ireland and internationally to provide evidence-based and robust decarbonisation insights for climate policy discussions.

@@

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%. \cite{EurostatShareSurveyilc_lvho50a}. The low building thermal efficiency is highlighted in Fig.\ref{fig:BERAge} and home heating oil accounts for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport} .

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}. 
In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

@@

The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock. The BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

The energy-related GHG emissions is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture. Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.

@@

\subsubsection{Heat}
Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually growing due to grants, awareness and ease of upgrades for homeowners. In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology. As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.  The energy efficiency first approach in how Ireland aims to 
decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
\end{itemize}
%PICKUPHERE
This report is a study of potential design options for a Renewable Heat Incentive (RHI)

-  critical for the development of renewable gas projects in Ireland.European Commission’s decision on State aid for the Support Scheme for Renewable Heat (SA.50807).
GNI Vision 2050 - 20\% renewable gas by 2030
While the electricity network reaches all residential dwellings, The gas network reaches 40\% or 680,000 dwellings in towns and cities. The 2030 gas decarbonisation targets, set by GNI  
and  with biomethance and renewable gas targets for 2030, X\%. The district heating networks are new and currently only two in Dublin which reach about XX houses, 2030 target X%. large amount of data centres. Other future heating fuel options such as hydrogen, HVO, solar may also play a part. 
Gas Networks Ireland (GNI) has a strategic plan to achieve 20% Renewable Gas on the network by 2030 
Further investigate decarbonisation options such as hydrogen vehicles, biomethane and AD
substitutes for natural gas - See \ref{fig:CAP2019:MACC} shows biomethane as the most expensive options..

\subsection{TIMES Ireland Model (TIM)}
\label{Intro:TIM}
\par
TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets. The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data. The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM . TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing. The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps. TIM will also be open platform to help improve the model and provide transparency. TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy. The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM. for these reasons TIM residential sector is a state of the art modelling platform. 
%GRAPH OF IRELAND DECARBONISATION TARGETS  VS OTHER COUNTRIES - DIFFERENT INTRASTATE and RESOURCE AT BIDENS ENERGY SUBMIT - DENMAKR TOP 2018, 
	ESOM can be used to start the conversation on..
Look at other ESOM, brief explain…
Other residential sectors..
CAP2019 - Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\section{Literature Review}
\label{LitR}
``How should Ireland set GHG targets in the residential sector?” 
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

%Lorem ipsum dolor sit amet, consectetur adipiscing \citep{Fabioetal2013} elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud \citet{Blondeletal2008} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit \citep{Blondeletal2008,FabricioLiang2013} anim id est laborum.

%Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum see appendix~\ref{sec:sample:appendix}.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\section{Acknowledgements}
\label{sec:Acknowledgements}
%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]

>>>>

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}
\label{Intro:Climate Targerts}
\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) which complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emission 2030 target would change from 30\% to 45\%. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

@@

Ireland is a unique EU member state, located on the peripherals of Europe, with a the second highest GDP per capita \cite{EurostatReal} (Higher GDP per capita translates to higher non-ETS reduction target), the largest proportion of agriculture GHG emissions at 34\% in 2018  \cite{EPA2020Irelands1990-2018}, and a high dependency on fossil fuel imports. Ireland has been afforded high non-ETS flexibility allowances under (EU) 2020/2126, to compensate for these barriers, but Ireland still has ambitious national climate targets. To put Ireland's climate targets in perspective Fig.\ref{fig:GHG2030AGR} compares different countries 2030 climate target with a 2018 base year and share of agriculture GHG emissions. Ireland has ambitious targets, despite distinct barriers. The Climate Change Advisory Council (CCAC) will rely of some modelling tools to provide insights and set sectoral five year carbon budgets.

@@

\subsection{Energy System Optimisation Models}
\label{Intro:ESOM}
\par % Paragraph 2 - ESOM
A Energy System Optimisation Model (ESOM) can optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex economic sectoral interconnections within each time period. The ESOM described in this paper is The Integrated MARKAL-EFOM System (TIMES) modelling framework, which has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA). TIMES is a ``hybrid" ESOM, which combines high resolution ``bottom-up" technological detail and ``top-down" macroeconomic variables covering all energy sectors.
\par In addition to TIMES there are other ESOM frameworks available such as EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}. However the expertise among energy policy researchers in Ireland combined with the global network of ETSAP modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. \par Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
While TIMES models can cover all the energy-related sectors, it can also be used to explore the residential sector in detail. Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}. There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets. Therefore, there is a need for a new TIMES model to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.

@@

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}. Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.  
In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

@@

The 2021 building regulations in 2021 require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock. The BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture. Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.

@@

\subsubsection{Heat}
%Gas Network Ireland (GNI) anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020ENERGYReport}.
Sustainable Energy Authority of Ireland (SEAI) have been played a leading role in the uptake of building fabric upgrades and heat pumps installations in existing homes. SEAI provide numerous grants to homeowners, increase public awareness of the benefits and are working on simplifying the process. In parallel, new dwellings must achieve a minimum of 20\% renewable energy, the ambient heat used in heat pumps is included as a renewable energy, often heat pumps are the preferred heating technology despite the high capital costs. As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.  The energy efficiency first approach in how Ireland aims to decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
\end{itemize}

\subsubsection{Gas \& District Heating}

In a bid not to be left behind in a ``net-zero" future, Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019AIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. GNI plans to met ``net-zero" in 2050 by using 50\% Carbon Capture and Storage (CCS), 37\% biomethane and about 13\% Hydrogen\cite{Ervia2019AIreland}. The gas network currently reaches 40\% or 680,000 dwellings in urban areas, but with the ``Compact Growth" as mentioned in section \ref{Intro:Policy} a higher share of building may have access to the gas network in the future. Increasing biomethane from an agricultural source in the gas network is the most expensive option in CAP2019's MACC shown in Fig.\ref{fig:CAP2019:MACC}.

Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland} \cite{Harney2020DeterminingIreland}
In Ireland,  DH networks are a new, despite being used for decade across central Europe. CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030, where the  Research in Ireland h
IrDEA Up to 57\% feasible - same as Denmark DH
37% of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80% 2050 reduction 
potential for 37% district heating nationally (Connollyet al., 2013a).
HVO, solar may also play a part.

\section{Literature Review}
\label{LitR}

``How should Ireland set GHG targets in the residential sector?” 
Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\subsection{Modelling Approach}
\label{LitR:Approach}
\subsection{Supply}
\label{LitR:Supply}
\subsection{Demand}
\label{LitR:Demand}
\subsection{Assumptions}
\label{LitR:Assumptions}
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

TIM was built to help Ireland provide a pathway to achieve 2030 and 2050 targets. The residential sector plays a key part of this and it was built from the SEAI 2018 energy balance, the BER database and CSO data. The data filtration process used in the residential sector, derived a lot of conversation of the BER database, a group know as Ireland Energy Wiki by the National Retrofitting Modelling Group grew from building the residential sector, questions about the filters, default values, assumptions around thermal equations (EC..) and uncertainties were used to improve the data before it was used in TIM . TIM has many internal project deadines, which required extensive meetings about the data, constraints and results as the modelling was progressing. The data and results were presented to 30 external reviewers to provide feedback, this process proved invaluable having different expertise with different presepectives enabled TIM to fill in gaps. TIM will also be open platform to help improve the model and provide transparency. TIM’s residential sector was built in parallel with LEAP’s residential sector, so that residential sector can easily simulate policy. The demand for results from TIM has increased significantly with increased government ambition, TIM will provide input into CAP 2021 and Ireland’s carbon budgets, working with DECC and the CCAC respectively will also provide feedback and improve the robustness of TIM. for these reasons TIM residential sector is a state of the art modelling platform. 
\section{Results}
\label{Results}
\subsection{Trajectories}
\label{LitR:Traj}
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\section{Conclusion}
\label{Conclusion}
\section{Discussion}
\label{Discussion}
\section{Acknowledgements}
\label{sec:Acknowledgements}
%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
